Automatic conversion of colloquial Finnish to standard Finnish

### 2015-03-28 2015-nodalida-finnormal/finnormal2015.tex
deanonymise
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59506 72bbbca6-d526-0410-a7d9-f06f51895060
changes suggested by reviewers
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@59547 72bbbca6-d526-0410-a7d9-f06f51895060

\author{}
%
%Francis M. Tyers\\
%  UiT Norgga \'{a}rktala\v{s} universitehta, \\
%  N-9015 Norway \\
%  {\tt francis.tyers@uit.no} \\\And
%  Inari Listenmaa\\
%  Chalmers tekniska h\"{o}gskola, \\
%  Sweden \\
%  {\tt inari@chalmers.se} \\}
%
\date{}

\begin{document}
\maketitle
\begin{abstract}
  This paper presents an unsupervised method for converting between colloquial Finnish
  and standard Finnish. The method relies upon a small number of orthographical rules
  combined with a large language model of standard Finnish for ranking the possible 
  conversions. Aside from this contribution, the paper also presents an evaluation
  corpus consisting of aligned sentences in colloquial Finnish, orthographically-standardised
  colloquial Finnish and standard Finnish. The methods we present outperforms the baseline
  of simply treating colloquial Finnish as standard Finnish and offers promise for the adaptation
  of language-technology tools created for standard Finnish to colloquial Finnish. To this end
  the paper also presents preliminary results which show promise for using normalisation in 
  the machine translation task.
\end{abstract}

@@

Most language technology tools are designed or trained based on standard language 
forms, where they exist. The application of these tools to non-standard
language can cause a substantial decrease in quality. For example in machine translation,
parsing and part-of-speech tagging \cite{eisenstein2013}. Non-standard language can 
have different orthographic conventions, along with different morphology, syntax and stylistics.

@@

There are a number of areas of research related to the task of text normalisation. Text
proofing tools, such as spelling and grammar checkers \cite{kukich1992} can be used to encourage adherence
to particular orthographic or grammatical norms. Accent and diacritic restoration --- for example in \newcite{scannell2011} --- is similar
in that it aims to bring text closer to standard orthography in order to facilitate treatment by 
automatic tools. Another related area is machine translation between different written norms of 
the same language. For example between Norwegian Bokm\aa{}l and Norwegian Nynorsk \cite{unhammer2009}.

@@

% Colloquial Finnish differs from standard Finnish in some key features:

% Morphology shows dialectal variation, shortening and merging of morphemes, e.g. \emph{en min\"{a}} `not I' $\rightarrow$ \emph{emm\"{a}} (`I don't').

% Syntax: due to the spontaneous nature of colloquial language, syntactic structures are common for spoken language. 
%Prototypically, colloquial language is spoken and spontaneous, and standard is written, planned and structured.

@@

For the normalised version,\footnote{The normalised version is converted orthographically and lexically, but not syntactically or stylistically.} we changed only morphology and vocabulary. On the lexical level, the word \emph{seiska} `number 7' is colloquial style, and in the standard translation it is replaced by the ordinal \emph{seitsem\"{a}s} `seventh'.  Other changes in the normalised version are common morphological or phonological, phenomena, such as reducing the diphthong in \emph{vain} $\rightarrow$ \emph{vaan} or shortening in  \emph{ett\"{a}} $\rightarrow$ \emph{et}. The original sentence and the normalised translation are shown below, aligned word by word.

@@

\ex
\begingl
\gla p\"{a}\"{a}ttelin seitsem\"{a}nnenkin fonotaksista //
\glb I.deduced seventh-{\sc also} phonotactics.{\sc ela} //
\glft `I deduced also the seventh from phonotactics' //
\endgl
\xe

% ``Yleens\"{a} kirjakieli ymm\"{a}rret\"{a}\"{a}n samaksi kuin yleiskieli: kielenhuollon suositusten mukaiseksi standardoiduksi kieleksi, jonka sanasto on yleistajuista. Puhekieli puolestaan miellet\"{a}\"{a}n kirjakielt\"{a} arkisemmaksi ja tyylilt\"{a}\"{a}n vapaamuotoisemmaksi kielimuodoksi, jonka ei tarvitse noudattaa kirjakielen s\"{a}\"{a}ntöj\"{a} (esim. mie, h\"{a}nen koira, me kirjotetaan vs. min\"{a}, h\"{a}nen koiransa, me kirjoitamme).''

@@

\begin{table}
  \centering
 \begin{tabular}{|l|r|r|r|}
    \hline
    \multirow{2}{*}{\textbf{Section}} & \multicolumn{3}{|c|}{\textbf{Tokens}} \\\cline{2-4}
                                      & \texttt{train} & \texttt{test} & \texttt{dev} \\
    \hline
    Colloquial                           & 5,103 & 1,012 & 1,003 \\ 
    Normalised                           & 5,105 & 1,012 & 1,003 \\ 
    Standardised                         & 4,982 & 991 & 1,000 \\
    \hline
 \end{tabular} 
  \caption{Statistics on sentences from the parallel corpus.}
  \label{table:corpsize}
\end{table}

@@

The corpus was split into three parts, testing, development and training. The testing
and development portions contain approximately 1,000 words each, with the remaining approximately 5,000 words for training
supervised models.\footnote{The corpus is split into 14 files of 500 words each. Files 01--02 
  were used for development; 03--04 for testing and 05--14 for training.} Table~\ref{table:corpsize} gives statistics on the number of words in each section.

@@

\subsection{Unsupervised normalisation}

For the unsupervised normalisation we applied a set of regular-expression based 
replace rules to the input text to produce all the possible candidate sentences 
in standard Finnish and then used a target-language model to rank the possible 
candidates. The candidate with the highest rank was selected as the normalised sentence.
For the target-language model we used the Finnish side of the English--Finnish EuroParl
parallel corpus \cite{koehn2005}.

We developed two sets of rules for the unsupervised normalisation approach:

@@

Table~\ref{table:unsup-trace} gives an example trace of the system on a simple sentence
using three replace rules.

    \hline
    \textbf{Output:} & \multicolumn{3}{l|}{Min\"{a} olen Tommin kanssa} \\
    \hline
  \end{tabular}
  \caption{Example trace of the unsupervised normalisation method. Rules are applied in order to each of 
     the possible candidate translations in turn. The candidates are then ranked using an $n$-gram language model 
     of standard Finnish and either an $n$-best list
     or the best candidate is output.}
  \label{table:unsup-trace}
\end{table*}

@@

The target-language model corpus, trained using KenLM \cite{heafield2011}, used was the same as in the unsupervised experiments.

@@

\begin{table}
  \centering
  \begin{tabular}{|l|r|r|r|}
     \hline
    \textbf{System} & \textbf{PER} & \textbf{WER} & \textbf{BLEU} \\
     \hline
     Baseline & 46.12 & 48.04 & 26.31 \\
     \hline
     rules-1 & 38.27 & 41.19 & 32.65 \\
     rules-2 & \textbf{38.17} & \textbf{35.25} & \textbf{36.41} \\
     rules-c & 36.56 & 39.68 & 34.68 \\
     \hline
     {\sc cbmt}-cn & 43.09 & 48.34 & 33.55 \\
     {\sc cbmt}-cs & 46.22 & 52.27 & 29.21 \\
     {\sc pbmt}-cn & \textbf{28.05} & \textbf{35.42} & \textbf{48.37} \\
     {\sc pbmt}-cs & 27.95 & 36.13 & 46.76 \\
     \hline
  \end{tabular}
  \caption{Results for the normalisation task. The system \texttt{rules-c} is the combination
     of the rules in \texttt{rules-1} and \texttt{rules-2}. The figures in bold are the 
     best results for unsupervised and supervised methods.}
  \label{table:results-norm}
\end{table}

@@

Our unsupervised method performs similarly to the character-based machine translation systems. Both 
the character-based systems and the unsupervised system achieve around half of the performance of the 
supervised phrase-based system.

Out of the unsupervised systems, the set of rules which was created by examining the development corpus
outperforms both the set of rules from the grammar and the combined rules.

@@

As for the methods, we would like to follow \newcite{scannell2014} in incorporating
`translation' probabilities into our unsupervised normalisation model. Our current model relies 
exclusively on the target-language model probability, however some rules may be more
reliable or probable than others.

@@

\textit{Removed for review}
%Joonas Kylm\"{a}l\"{a}

>>>>

\author{ Inari Listenmaa\\
  Chalmers tekniska h\"{o}gskola, \\
  Sweden \\
  {\tt inari@chalmers.se} \\\And
 Francis M. Tyers\\
  HSL-fakultehta, \\
  UiT Norgga \'{a}rktala\v{s} universitehta, \\
  N-9015 Norway \\
  {\tt francis.tyers@uit.no} }

\date{}

\begin{document}
\maketitle
\begin{abstract}
  This paper presents a rule-based method for converting between colloquial Finnish
  and standard Finnish. The method relies upon a small number of orthographical rules
  combined with a large language model of standard Finnish for ranking the possible 
  conversions. Aside from this contribution, the paper also presents an evaluation
  corpus consisting of aligned sentences in colloquial Finnish, orthographically-standardised
  colloquial Finnish and standard Finnish. The method we present outperforms the baseline
  of simply treating colloquial Finnish as standard Finnish, but is outperformed by a phrase-based 
 MT system trained by the evaluation corpus.
% Nevertheless, it offers promise for the adaptation of language-technology tools created for standard Finnish to colloquial Finnish. To this end
  The paper also presents preliminary results which show promise for using normalisation in 
  the machine translation task.
\end{abstract}

@@

Most language technology tools are designed or trained based on standard language 
forms, where they exist. The application of these tools to non-standard
language can cause a substantial decrease in quality for example in machine translation,
parsing and part-of-speech tagging \cite{eisenstein2013}. Non-standard language can 
have different orthographic conventions, along with different morphology, syntax and stylistics.

@@

There are a number of areas of research related to the task of text normalisation. Text
proofing tools, such as spelling and grammar checkers \cite{kukich1992} can be used to encourage adherence
to particular orthographic or grammatical norms. Accent and diacritic restoration --- for example in \newcite{scannell2011} --- is similar
in that it aims to bring text closer to standard orthography in order to facilitate treatment by 
automatic tools. Another related area is machine translation between different written norms of 
the same language, for example between Norwegian Bokm\aa{}l and Norwegian Nynorsk \cite{unhammer2009}.

@@

@@

For the normalised version,\footnote{The normalised version is converted orthographically and lexically, but not syntactically or stylistically.} we changed only morphology and vocabulary. On the lexical level, the word \emph{seiska} `number 7' is colloquial style, and in the standard translation it is replaced by the ordinal \emph{seitsem\"{a}s} `seventh'.  Other changes in the normalised version target common morphological or phonological phenomena, such as restoring the reduced diphthong in \emph{vaan} $\rightarrow$ \emph{vain}. The original sentence and the normalised translation are shown below, aligned word by word.

@@

\ex
\begingl
\gla p\"{a}\"{a}ttelin seitsem\"{a}nnenkin fonotaksista //
\glb I.deduced seventh-{\sc also} phonotactics.{\sc ela} //
\glft `I deduced the seventh also from phonotactics' //
\endgl
\xe

@@

\begin{table}
  \centering
 \begin{tabular}{|l|r|r|r|}
    \hline
    \multirow{2}{*}{\textbf{Section}} & \multicolumn{3}{|c|}{\textbf{Tokens}} \\\cline{2-4}
                                      & \texttt{dev} & \texttt{test} & \texttt{train} \\
    \hline
    Colloquial                           & 1,003 & 1,012 & 5,103 \\ 
    Normalised                           & 1,003 & 1,012 & 5,103 \\ 
    Standardised                         & 1,000 & 991 & 4,982 \\
    \hline
 \end{tabular} 
  \caption{Statistics on sentences from the parallel corpus.}
  \label{table:corpsize}
\end{table}

@@

The corpus was split into three parts, development, testing and training. The development and testing portions contain approximately 1,000 words each, with the remaining approximately 5,000 words for training
phrase-based and character-based models.\footnote{The corpus is split into 14 files of 500 words each. Files 01--02 
  were used for development; 03--04 for testing and 05--14 for training.} Table~\ref{table:corpsize} gives statistics on the number of words in each section.

    \hline
    \textbf{Output:} & \multicolumn{3}{l|}{Min\"{a} olen Tommin kanssa} \\
    \hline
  \end{tabular}
  \caption{Example trace of the normalisation method. Rules are applied in order to each of 
     the possible candidate translations in turn. The candidates are then ranked using an $n$-gram language model 
     of standard Finnish and either an $n$-best list
     or the best candidate is output.}
  \label{table:normal-trace}
\end{table*}
  \newpage

@@

\subsection{Rule-based normalisation}

For the rule-based normalisation we applied a set of regular-expression based 
replace rules to the input text to produce all the possible candidate sentences 
in standard Finnish and then used a target-language model to rank the possible 
candidates. The candidate with the highest rank was selected as the normalised sentence.
For the target-language model we used the Finnish side of the English--Finnish EuroParl
parallel corpus \cite{koehn2005}.

We developed two sets of rules:

@@

Table~\ref{table:normal-trace} gives an example trace of the system on a simple sentence
using three replace rules.

@@

The target-language model corpus, trained using KenLM \cite{heafield2011}, used was the same as in the rule-based experiments.

@@

\begin{table}
  \centering
  \begin{tabular}{|l|r|r|r|}
     \hline
    \textbf{System} & \textbf{PER} & \textbf{WER} & \textbf{BLEU} \\
     \hline
     Baseline & 46.12 & 48.04 & 26.31 \\
     \hline
     rules-1 & 38.27 & 41.19 & 32.65 \\
     rules-2 & \textbf{38.17} & \textbf{35.25} & \textbf{36.41} \\
     rules-c & 36.56 & 39.68 & 34.68 \\
     \hline
     {\sc cbmt}-cn & 43.09 & 48.34 & 33.55 \\
     {\sc cbmt}-cs & 46.22 & 52.27 & 29.21 \\
     {\sc pbmt}-cn & \textbf{28.05} & \textbf{35.42} & \textbf{48.37} \\
     {\sc pbmt}-cs & 27.95 & 36.13 & 46.76 \\
     \hline
  \end{tabular}
  \caption{Results for the normalisation task. The system \texttt{rules-c} is the combination
     of the rules in \texttt{rules-1} and \texttt{rules-2}. The figures in bold are the 
     best results for rule-based and SMT methods.}
  \label{table:results-norm}
\end{table}

@@

Our rule-based method performs similarly to the character-based machine translation systems. Both 
the character-based systems and the rule-based system achieve around half of the performance of the phrase-based system.

Out of the rule-based systems, the set of rules which was created by examining the development corpus
outperforms both the set of rules from the grammar and the combined rules.

@@

As for the methods, we would like to follow \newcite{scannell2014} in incorporating
`translation' probabilities into our rule-based normalisation model. Our current model relies 
exclusively on the target-language model probability, however some rules may be more
reliable or probable than others.

@@

Joonas Kylm\"{a}l\"{a}
